### 2016-02-13 Manuscript/BiologyLetters/response.tex

% *Please upload your original figure files as eps, tiff or jpeg files, rather than PDFs
%TG: TO DO (easy)

%%%%%%%%%%%%%%%
% Reviewer 1
%%%%%%%%%%%%%%%

\section{Reviewer 1:}
\begin{enumerate}
\item{\textcolor{blue}{In ESM 1, p. 4, the authors explain their methods of collecting morphological data from the literature. Why select only the first 10 results per search term? How do you know that relevant papers wouldn't come up later in the search? I need evidence to support this cutoff; as it stands, I'm concerned that you're missing some relevant material.}}
% NC: It's not just the first 10 results per search term is it? I thought you went through the first 10 pages of results? Clarify this. TG: No I collected the 20 first results. It's mentioned down there.

As shown in ESM1 Figure 1, the number of OTUs with morphological cladistic data extracted from the Google Scholar searches seems to plateau really quickly (i.e. before the first 100 matches).
Additionally, as Graeme Lloyd mentions, ``older studies tend to be smaller (due to computing limitations) and to commonly reoccur anyway due to data set reuse'' (Reviewer 4, comment 1).
Following these two observations, we think that including the 20 most recent results for each order (and higher taxonomic level) was a good balance between the time spent looking for papers, the number of OTUs extracted and the repeatability of the study.
We are aware however that we might have missed some specific studies (see Reviewer 2 comment 2) but are confident that our protocol gives a good overall picture of the data availability and that the three repositories used before the Google Scholar search (MorphoBank, Graeme Lloyd's and Ross Mounce's) contained most of the available data.
In fact, as mentioned in ESM 1, part 1 ``among the 660 papers [collected from the Google Scholar search], only 50 contained a total of 425 extra living OTUs''. % NC: How many OTUs were in the other 610 papers?

@@

\item{\textcolor{blue}{Similarly, I find the 2010 date cutoff troubling, because I'm not sure that the argument that these recent studies contain a fraction of the previously published characters and OTUs allows you to make a comprehensive statement about the availability of character data. What about the fraction that isn't included? This is a question that needs to be better addressed before you will convince readers that this is really all the data that is out there.}}

We chose this cutoff date to filter our analysis towards accessible data only (see Reviewer 2 comment 2).
Additionally, we observed that in newer studies (i.e. post 2010), living taxa with anatomical data were rarely added and usually reused from former studies (with the occasional addition of extra characters and extra living taxa).
We modified the sentence in the ESM1 part 1:

@@

\item{\textcolor{blue}{P3, L78: ``since'' should be ``because.'' Since should be used for time, not causation.}}
We replaced ``since'' with ``because'' throughout the manuscript as appropriate.

\item{\textcolor{blue}{P3, L85: ``not-applicable'' should not be hyphenated.}}
We removed the hyphen.

\item{\textcolor{blue}{P3, L120: How much of a difference does the 25\% or less coverage make to the capacity of Total Evidence phylogenetic analyses to resolve relationships? Does your simulation study have a way of quantifying or describing the probability of getting the ``right'' answer with this little data? It would also help if you could show in your data table what the clades were that had better than 75\% coverage. It's not immediately evident from reading Table 1.}}

The relationship between the living species coverage and the Total Evidence phylogeny topology is hard to quantify precisely and depends on both the metric used for measuring topological differences and the method used to infer the tree.
Our simulations showed that topology tends to get closer to random as morphological data is removed but that in the case of missing data for living species, the other parameters we tested (missing data in the fossil record and number of morphological characters) had no significant effect (i.e. when a certain number of living taxa are missing, the other parameters do not have any extra influence).
In our simulations we tested five different levels of missing living taxa (0\%, 10\%, 25\%, 50\% and 75\% missing living taxa) and we observed that the three first levels had no significant negative effect on topology in relation with the amount of missing data in the fossil record and the number of morphological characters but that the fifth level (75\% of missing living taxa, i.e. 25\% coverage) had a strong negative effect.

Also, we highlighted the different levels of coverage in Table 1 as follow: at any taxonomic level, orders with less than 25\% coverage are highlighted in blue (or dark grey in the black and white version) and orders with more than 75\% coverage are highlighted in orange (or light grey in the black and white version).

@@

\end{enumerate}

%%%%%%%%%%%%%%%
% Reviewer 2 (Graham Slater)
%%%%%%%%%%%%%%%

\section{Reviewer 2 (Graham Slater):}
\begin{enumerate}
\item{\textcolor{blue}{The authors state on page 3 lines 82-90 that they avoided matrices with ``few characters'' due to lower probability of overlapping characters with other matrices and so only selected matrices containing >100 characters. First, a lack of overlapping characters doesn’t seem justifiable, at least to my mind, as a reason to exclude a matrix if the ultimate goal is to infer the placement of extinct taxa relative to extant taxa. Antler ramifications, to use the authors' example, are likely to be a very useful character when attempting to resolve relationships among living and extinct artiodactyls, even if they cannot be coded for monotremes. To flip the argument, a matrix of 200 + characters for mysticetes is similarly likely to contain many non-applicable characters from the perspective of talpids. I do understand that the authors are thinking downstream in terms of super matrices for higher level analyses, but smaller matrices are going to be just as relevant for reaching this goal. You don't have much space in the ms I understand, but surely this needs to be addressed?!}}

We completely see the reviewer's point and have removed this threshold and replaced it with the analysis without any minimal number of morphological characters.
We also removed the paragraph about the character threshold analysis and updated the results of figure 1 and table 1 to contain all living OTUs; we removed the redundant ESM 2 and we updated the figures.

@@

\item{\textcolor{blue}{Along these lines, a 100 character limit may seem rational but this is actually, from a practical perspective, very large. molecular matrices of this size would, of course, be laughed at but many of the most robust phylogenetic studies I'm familiar with come in well under this size. I suspect, but cannot be sure, that this partially explains the poor species level coverage the authors find; higher level studies can readily accumulate large numbers of characters because vastly different morphologies are being compared, often using families or genera as exemplars, leading to an abundance of easily defined characters. Species level studies, on the other, are often focused within families or genera to make them more manageable, but here characters are quickly exhausted. To give some random examples that I'm familiar with, Salles (1992) species-level study of Felidae (all species) was based on 44 characters, Wang and Carranza- Casta\~{n}eda's 2008 study of Mephitidae (\textasciitilde50\% of species but all genera) was based on 38 characters, and Bryant et al's 1993 study of Mustelid genera was of 46 characters. These are all important studies, but do not feature here because of the 100 character limit. At the very least I think this results in a downplaying of previous phylogenetic efforts by morphologists in the results. At worst it misrepresents the state of affairs dramatically. Is 100 really the minimum number you want to go with at all phylogenetic scales?}}

Again we completely see the reviewer's point and have modified our manuscript as follows.
By removing the threshold (see above) we now included the taxa coded in Wang and Carranza-Casta\~{n}eda 2008 (matrix reference GS2014-WCC) and other matrices excluded in the former analysis increasing the number of Carnivora from 42 species (14\% of total number of species) to 76 (27\%).
However, the two other matrices mentioned by reviewer 2 (Salles 1992 and Bryant et al. 1993) were not included in the new analysis.
This is due to two main reasons: firstly, these two matrices were not available on any of the repositories (either Graeme Lloyd's, Ross Mounce's or MorphoBank's), and secondly, they didn't appear in our additional Google Scholar searches because they were published before 2010.
Following this reviewer's comment as well as reviewer 4's we added a paragraph in the discussion to clearly state that we are aware of data sampling biases:

``It is worth noting, however, that our analysis did not include all the matrices containing anatomical characters ever published.
In fact, our data collection procedure focused on including studies that provided matrices easily accessible, i.e. we did specifically not include any matrices that were only available in paper format (e.g. printed in books), non-reusable format (e.g. an image of the matrix) %TG: yes... There were quiet a few of them!
or/and matrices available only upon request (e.g. by emailing the authors).
Such matrices are likely to not be included in Total Evidence analysis due to the difficulty of obtaining them. % NC: I think this last line is a bit dangerous. Anyone making a proper phylogeny would include these, especially for a smaller group. It's what people like Graham spend years doing. Perhaps rephrase: "The difficulty of obtaining such matrices makes them less likely to be included in Total Evidence analyses, especially where the number of taxa to include in the phylogeny is high."
'' p.@ lines@@.

@@

\item{\textcolor{blue}{I finally am a little skeptical of the sampling done. Looking at Figure 1B, again the clade I'm most familiar with, representation for carnivorans appears very poor. Notwithstanding the issues discussed above regarding matrices with < 100 characters, there are definitely matrices out there that fulfill the authors' requirements that are not represented. For example, a 349 character matrix with near complete sampling at the species is available for Viverridae, Prionodontidae, and Eupleridae in Gaubert et al (2005 Syst Biol.). Going through each clade in the supplement seems beyond the scope of reviewing duties, but I am concerned that if this one matrix can be missed others may have been missed too. I think a reexamination of the literature may be warranted using taxon specific search terms.}}

We dealt with this comment by removing the matrix size threshold and adding a paragraph to the discussion explaining that some matrices might not be included in this study (see reviewer's 2 comments 1 and 2).
Concerning the Gaubert et al. 2005's study specifically, this study unfortunately did not contain any easily available matrix.
In fact, the matrix mentioned as Appendix 2 in the paper is not present in the article and neither of the TreeBase accession numbers (``S1255'', ``M2193'') nor the provided URL (``www.systematicbiology.org'') links to the matrix.
% TG: I don't know if that's a good justification but basically, it's impossible to get this specific matrix.
% NC: Hmmm it's tricky as I said above. If you were really building a tree you would find this kind of data somehow. Also wouldn't this not appear because you used the 2010 cutoff? I think tbh Graham has some good points - you did this the quick and dirty way rather than the thorough way someone would do it if they were constructing a tree.

@@

\item{\textcolor{blue}{One interesting aside that the authors didn't consider, despite the data being there, is how phylogenetic ``effort'' relates to diversity of the order under study.}}

This is a great suggestion, thanks!
We added this analysis to the ESM 2 (Supplementary results - 1) showing, as suggested, that mammalian morphological phylogeneticists have focused their energies on resolving the higher level relationships among mammalian lineages.
We also added a sentence mentioning these results in the manuscript:

``It is also encouraging to note that the sampling effort is consistent with diversity at each taxonomic level and there is therefore no direct bias in sampling at any taxonomic level (ESM 2, Figure 1).
'' p.@ lines@@

% NC: OK this is not what Graham meant. I think what he is referring to is that when you have more people working on a group (effort) you are likely to get more taxa with coded data. It's like in taxonomy where groups with more taxonomists (eg primates) have more species than groups with fewer (eg rodents). I think what he'd want is a regression of number of papers (effort) against percentage of coded species/OTUs with each order being a datapoint.

@@

\end{enumerate}

%%%%%%%%%%%%%%%
% Reviewer 3 (Peter Wagner)
%%%%%%%%%%%%%%%

\section{Reviewer 3 (Peter Wagner):}
\begin{enumerate}
\item{\textcolor{blue}{My biggest criticism is a semantic one: Morphological data $\neq$ cladistic data.}}
We agree with this reviewers semantinc point and changed ``cladistic'' and ``morphological'' data to ``anatomical characters'' throughout the manuscript.
% NC: all he is saying is cladistic is not morphological and shouldn't be used as proxies for each other. And that really when we say morphological we mean phylogenetically coded anatomical/morphological characters. I've changed a few of these back in the manuscript to try and make it a bit easier to understand, as anatomical suggests something else I think.

@@

\item{\textcolor{blue}{The real issue here is the availability of phylogenetic character coding for anatomical data.
[...] Thus, the title really should be ``Assessment of available anatomical characters for phylogenetic analysis among living mammals'' or something like that.}}
We changed the title as the reviewer suggested.

\item{\textcolor{blue}{what we really need is to focus on the tooth and skull characters that typify the fossil record.
[...] Still, several papers estimate that the sampling of mammal species (based on cranial characters include teeth) actually is pretty darned good [...].
Thus, we might be able to richly populate FBD and tip-dating analyses with a LOT of extinct species.}}

We totally agree with this comment.
In fact, among the 286 matrices used in this analysis, we found 5228 unique OTUs including 3627 fossil ones which is pretty impressive knowing the amount of work required to collect data for a single specimen!
However, in this study, we focused only on the number of *living* taxa (1601 unique OTUs containing 847 species) as a way to solve the problems associated with the lack of living taxa on tree topology described in Guillerme \& Cooper 2016 (essentially that if you don't have overlapping coded anatomical data for your living species then there is a low likelihood of correctly placing the fossil taxa no matter how good the sampling is for those species).

@@

\item{\textcolor{blue}{Finally, I would add: who the heck submits a paper for a special volume 6 weeks early?!?!? You make the rest of us look bad!}}
This paper formed part of my PhD thesis that I submitted in September which partially explains the swift submission!
Sorry for making the rest of the authors submitting to the issue look bad.

\end{enumerate}

\subsection{Reviewer 3 specific comments:}
\begin{enumerate}
\item{\textcolor{blue}{Lines 25-26: ``We suggest that increased morphological data collection efforts for living taxa are needed to produce accurate Total Evidence phylogenies.''
In a way, this is more than ``total evidence''.
The initial total evidence paradigm was simply to use every character you could (molecular, morphological, even behavioral).
We now are talking about tip-dating and fossilized birth-death (FBD) approaches (e.g., Heath et al. 2014 PNAS 111:E2957) that want to use all of the available data.
After all, one can do a total evidence analysis with just node-dating techniques rather than with tip-dating, and one can do it without worrying about branch length priors.
However: if we want to use all of the data to calibrate branch lengths for all taxa (fossil and extant), then we need ``total evidence'' to do tip-dating and FBD properly).
As tip-dating is the main theme of this issue, that should be linked to the total evidence explicitly here and elsewhere in the manuscript.}}

We are aware of the difference between the ``Total Evidence method'' (i.e. a way to combine data in a matrix like supermatrix) and ``tip-dating method'' (i.e. a phylogenetic dating method, like node dating).
Both methods can (and are often) be used together as in Ronquist et al 2012 (Systematics Biology) but we are aware that they can also be used really well separately.
However, this study focuses \textit{indeed} on the ``Total Evidence method'' (i.e. a way to combine molecular and discrete anatomical data together) and not on the ``tip-dating'' one because it attempts to quantify the problem of missing data in living taxa underlined in Guillerme \& Cooper 2016 (again, a paper focusing on the ``Total Evidence method'', not including ``tip-dating'' nor ``node dating'' analysis).
Nonetheless, we do understand the reviewer's point and added the mention of ``Total Evidence tip-dated phylogenies'' which is undoubtedly a great avenue of application of both methods.

We modified the sentence in the abstract as follows: ``We suggest that increased morphological data collection efforts for living taxa are needed to produce accurate Total Evidence tip-dated phylogenies.'' p@ lines@@

@@

\item{\textcolor{blue}{Lines 70-71: ``We downloaded all cladistic matrices containing any living and/or fossil mammal  taxa from three major public databases:''
As a side note and for future reference for the authors, a large number of phylogenetic data sets for mammals (and other metazoans) also are available at the Paleobiology Database.
(These were formerly at my website at the Field Museum, but that site obviously disappeared when I moved to the Smithsonian.)
Because of the nature of the projects for which I was using these datasets, most of them are species-level, but some are genus-level, too.}}

We thank the reviewer for this mention and we will be sure to use this database for future studies.
% NC: If you end up needing to redo the data collection it might be worth checking the PBDB...

\item{\textcolor{blue}{Lines 164-166: ``For example, a Carnivora fossil will be unable to branch in the Herpestidae, and will have more chance to randomly branch within Canidae (Figure 1B).'' 
This seems to be an awkward way to phrase this.
Moreover, this seems like an extension of the long-branch problem.
Basically, the issue really is that we do not have coded skeletal characters for extant herpestids.
Thus, it is not possible for coded skeletal characters from fossil herpestids to connect those fossil herpestids to the extant herpestids.
This is the opposite of Felsenstein's long branch: insofar as skeletal characters are concerned, extant herpestids have zero-length branches.
That in turn means that the fossil herpestids are going to be ``attracted'' to non-zero length branches elsewhere, with long branches in related groups such as canids being particularly good candidates because there is a higher chance of homoplasies between canids and herpesitds on the longer branches.
So, it will not be random: they probably will latch themselves onto the longest reasonably branch among the closest relatives for which anatomical data are present.}}

We agree with this comment and changed this to:
``For example, a Carnivora fossil will be unable to be placed in the Herpestidae clade because they have no anatomical characters available (i.e. extant Herpestidae have a zero-length branch for anatomical characters).
Such fossils, however, will have a higher probability of being placed on a branch that contains many anatomical characters (i.e. long branches for anatomical characters, e.g. Canidae or Ursidae (Figure 1B).''

@@

\end{enumerate}

%%%%%%%%%%%%%%%
% Reviewer 4 (Graeme Lloyd)
%%%%%%%%%%%%%%%

\section{Reviewer 4:}
\begin{enumerate}
\item{\textcolor{blue}{Reviewer 1's most substantial complaints (the year and page number cut offs for google scholar searches) seem at least partially valid.
However, in my experience older studies tend to be smaller (due to computing limitations) and to commonly reoccur anyway due to data set reuse.
The authors also already state that subsequent search results tended to yield no substantial additional data, and by the reviewer's own admission the assembled data seem to look correct.
Thus a recollecting of the data doesn't seem necessary (but see further comments below).}}

We developed the description of our data collection protocol in the ESM 1 (see Reviewer 1 comment 2) and underlined the caveats in the discussion (see Reviewer 2 comment 2).

### 2016-02-13 Manuscript/BiologyLetters/response.tex

\item{\textcolor{blue}{Reviewer 2 (Graham Slater)'s complaint are - in my view - the most substantial.
I agree that the >100 characters cut off is not as strongly justified as it might be.
However, as the authors have run their protocol without this threshold this is easily fixed by at least briefly mentioning how this affects their results in the main text (not relegating it to the ESM where no actual discussion is provided).
From my reading of the two tables only 11 of the orders seem to have below 25\% coverage at the species-level (not 22 as the authors state in the main text!) when the threshold is used, and this improves slightly to 10 for when the threshold is removed.
As this difference isn't dramatic it seems worth mentioning.
Reviewer 2 also notes that some important studies may be missed.
This seems valid (I am aware of the biases in Ross Mounce's data collection protocol, and obviously my own) and I am sure there are many more data sets that could be sampled.
In particular I am aware of a number of studies that can only be found in books and hence are not easily accessed even if they occur in a google scholar search.
I do not expect the authors to more thoroughly search the literature for all possible cladistic analyses of mammals, however, as this is a major task.
(I should know, it has been a long term project of mine for years.)
Nevertheless, it does deserve mention in the discussion as a potential bias in the results.
I don't think this undermines the study, but rather should be viewed as both a reason for optimism and an acknowledgement of a separate problem (i.e., access to cladistic data) which multiple publications already cover (e.g., see Ross Mounce's work).}}

We thank the reviewer for their practical stance on recollecting the dataset. We have added some comments to the manuscript to deal with some of the biases in our procedure and why some matirces will be missing.
We also removed the character threshold and specified what we meant by ``data accessibility'' in the discussion (see Reviewer 2 comment 2). See also our responses to the reviewer comments above.

### 2016-02-13 Manuscript/BiologyLetters/Missing_mammals-draft.tex
Dealt with comments

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------
\newpage
\begin{abstract}
\section{abstract} % NC: Just put this here as it then separates this for the word count

% NC: I think anatomical here confuses things as we mean coded anatomical characters. I've tried to clarify.  Also needed to clarify what you mean by "available". Also had to then cut some words to get it back under 200.
Analyses of living and fossil taxa are crucial for understanding biodiversity through time.
The Total Evidence method allows living and fossil taxa to be combined in phylogenies, using molecular data for living taxa and morphological data for living and fossil taxa.
With this method, substantial overlap of coded anatomical characters among living and fossil taxa is vital for accurately inferring topology.
However, although molecular data for living species are widely available, scientists generating morphological data mainly focus on fossils.
Therefore, there are few coded anatomical characters in living taxa, even in well-studied groups like mammals.

We investigated the number of coded anatomical characters available in phylogenetic matrices for living mammals and how these were phylogenetically distributed across orders.
22 of 28 mammalian orders have \textless 25\% species with available characters; this has implications for the accurate placement of fossils, although the issue is less pronounced at higher taxonomic levels. 
In most orders, species with available characters are randomly distributed across the phylogeny, which may reduce the impact of the problem.
We suggest that increased morphological data collection efforts for living taxa are needed to produce accurate Total Evidence phylogenies. 
% NC: I removed the tip dating bit here, I think yes peter is right in that we are in a special issue on that, BUT it is important for all TEM analyses so adding the tip dating bit here is daft.
\end{abstract}

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\newpage 
\section{Introduction}

There is an increasing consensus among biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \cite{slaterunifying2013,fritzdiversity2013}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \cite{ronquista2012}, combines molecular data from living taxa and morphological data %TG: I left "morphological data" here since it TEM it doesn't have to be only anatomical characters 
 from both living and fossil taxa in a supermatrix that can then be used with dating methods such as the tip-dating method (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014}), to produce a chronogram with living and fossil taxa at the tips. 
A downside of this method is that it requires molecular data for living taxa and morphological/anatomical data for both living and fossil taxa. %TG: same comment as above. I agree but have added anatomical to make it clear here what you mean when suddenly we switch to using anatomical.
Sections of this data can be difficult, or impossible, to collect for every taxon in the analysis.
For example, fossils rarely have molecular data and incomplete fossil preservation may reduce the number of anatomical characters available.
Additionally, it has become less common to collect anatomical characters for living taxa when molecular data is available (e.g. in \cite{slaterphylogenetic2013}, only 13\% of living taxa have coded anatomical characters).
Unfortunately this missing data can lead to errors in phylogenetic inference.
Simulations show that the ability of the Total Evidence method to recover the correct topology decreases when there is little overlap between coded anatomical characters in living and fossil taxa, and that the effect of missing data on topology is greatest when living taxa have few anatomical characters available \cite{GuillermeCooper}.
This is because (1) fossils will not be placed accurately within the correct clade if it contains no coded anatomical characters for living taxa; and (2) fossils have a higher probability of being placed within clades with more coded anatomical characters available for living taxa, regardless of whether this is the correct clade \cite{GuillermeCooper}.

The issues above highlight that it is crucial to have sufficient coded anatomical characters available for living taxa in a clade before using Total Evidence approaches.
However, it is unclear how many coded anatomical characters are actually available for living taxa, i.e. already coded from museum specimens and deposited in phylogenetic matrices accessible online, and how this data is distributed across clades. % TG: emphasize the "availibility" definition her?
Intuitively, most people assume this kind of data has already been collected, but empirical data suggest otherwise (e.g. in \cite{ronquista2012,slaterphylogenetic2013,beckancient2014}).
To investigate this further, we assess the number of available coded anatomical characters for living mammals to determine whether enough data exists to build reliable Total Evidence phylogenies.
We also determine whether the characters are phylogenetically overdispersed or clustered across mammalian orders.

%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------
% NC: I've done a bit of word cutting here as I think we had wordcount issues?
\section{Materials and Methods}
\subsection{Data collection and standardisation}
We downloaded all cladistic matrices %TG: I left "cladistic" here 'cause that's what the matrix is (even following Peter's definition)! % NC: Yep
 containing any living and/or fossil mammal taxa from three major public databases: MorphoBank (\url{morphobank.org} \cite{morphobank}), Graeme Lloyd's website (\url{graemetlloyd.com/matrmamm.html}) and Ross Mounce's GitHub repository (\url{github.com/rossmounce/cladistic-data}).
We also performed a systematic Google Scholar search for matrices that were not uploaded to these databases (see Electronic Supplementary Material (ESM) for details).
In total, we downloaded 286 matrices containing 5228 unique operational taxonomic units (OTUs). 
We used OTUs rather than species because entries in the matrices ranged from species to families. 
We standardised the taxonomy as described in the ESM and excluded OTUs that were not present in the phylogeny of \cite{BinindaEmonds} or the taxonomy of \cite{wilson2005mammal} to remove fossil species.

% Matrices with few characters are problematic when comparing available data among matrices because (1) they have less chance of having overlapping characters with other matrices \cite{wagner2000} and (2) they are more likely to contain specific characters that are not applicable% rev 1.2.4
%  across large clades (e.g. ``antler ramifications'' is a character that is only applicable to Cervidae \cite{Brazeau2011}).
% Therefore we selected only matrices containing \textgreater 100 characters for each OTU.
% This threshold was chosen to correspond with the number of characters used in \cite{GuillermeCooper} and \cite{harrisonamong-character2014}.
% Results of analyses with no threshold are available in the ESM. 
% After removing matrices with \textless 100 characters, we retained 1074 unique living OTUs from 126 matrices. % 1601 unique living OTUs for 286 matrices (no threshold)
This resulted in 1601 unique OTUs from 286 matrices.

\subsection{Data availability and distribution}
To assess the availability of coded anatomical characters for each mammalian order and across mammals, we calculated the percentage of OTUs with cladistic data at three different taxonomic levels: family, genus and species. %TG: changed cladistic datat to anatomical characters
We consider orders with \textless 25\% of living taxa with available anatomical characters as having low data coverage, and orders with \textgreater 75\% of living taxa with available anatomical characters as having high data coverage.

@@

\input{Table_results}

@@

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{figure1.pdf}
\caption{Phylogenetic distribution of species with available coded anatomical characters across two orders (A: Primates; B: Carnivora).
Blue branches indicate species with available coded anatomical characters.}
\label{Figure_example_coverage}
\end{figure}

@@

It is worth noting that our analysis did not include all the matrices containing coded anatomical characters ever published.
Instead of performing exhaustive searches, our data collection procedure focused on including studies that provided easily accessible matrices.
We specifically did not include matrices that were only available in books, in non-reusable formats (e.g. an image of the matrix that would require text-mining tools or recreating the matrix by hand), or matrices only available upon request.
The difficulty of obtaining such matrices makes them less likely to be included in Total Evidence analyses, especially where the number of taxa to include in the phylogeny is high and time is limited.

Despite the absence of good morphological/anatomical data coverage for living mammals, %TG: I think leaving cladistic here is OK no? % NC: Better to be consistent with rest of manuscript
 the Total Evidence method still seems to be the most promising way of combining living and fossil species in macroevolutionary analyses.
%It is also encouraging to note that sampling effort is consistent with diversity at each taxonomic level and there is therefore no direct bias in sampling at any taxonomic level (ESM 2, Figure 1).
% NC: Needs more explanation, why is this encouraging??? Maybe leave this as the paper is already over long.
Following the recommendations in \cite{GuillermeCooper}, we must code anatomical characters for as many living species as possible. 
Fortunately, mammal specimens are usually readily available in natural history collections, therefore, we propose increased effort into coding anatomical characters from living species, possibly by engaging in collaborative data collection projects.
Such efforts would be valuable not only to phylogeneticists, but also to any researcher focusing understanding macroevolutionary patterns and processes.

%Biology letters various stuff
\section{Ethics statement}
N/A
\section{Data accessibility statement}
All data and code are available on GitHub (\url{https://github.com/TGuillerme/Missing_living_mammals}).
\section{Authors' Contributions}
TG and NC designed the study. TG analysed the data. TG and NC wrote the the manuscript.
\section{Competing Interests}
We have no competing interests.
\section{Acknowledgments}
We thank David Bapst, Graeme Lloyd, Nick Matzke, April Wright, Graham Slater, Peter Wagner and one anonymous reviewer.
%TG: That's two times Graeme though...I've shortened it due to wordcount issues
\section{Funding statement}
This was funded by a European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant (proposal: 321696).

>>>>

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------
\newpage
\begin{abstract}
\section{abstract}

Analyses of living and fossil taxa are crucial for understanding biodiversity through time.
The Total Evidence method allows living and fossil taxa to be combined in phylogenies, using molecular data for living taxa and morphological data for living and fossil taxa.
With this method, substantial overlap of coded anatomical characters among living and fossil taxa is vital for accurately inferring topology.
However, although molecular data for living species are widely available, scientists generating morphological data mainly focus on fossils.
Therefore, there are few coded anatomical characters in living taxa, even in well-studied groups like mammals.

We investigated the number of coded anatomical characters available in phylogenetic matrices for living mammals and how these were phylogenetically distributed across orders.
11 of 28 mammalian orders have \textless 25\% species with available characters; this has implications for the accurate placement of fossils, although the issue is less pronounced at higher taxonomic levels. 
In most orders, species with available characters are randomly distributed across the phylogeny, which may reduce the impact of the problem.
We suggest that increased morphological data collection efforts for living taxa are needed to produce accurate Total Evidence phylogenies. 
\end{abstract}

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\newpage 
\section{Introduction}
There is an increasing consensus among biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \cite{slaterunifying2013,fritzdiversity2013}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \cite{ronquista2012}, ccombines molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix that can then be used with the tip-dating method (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014}), producing a chronogram with living and fossil taxa at the tips.
A downside of this method is that it requires molecular data for living taxa and morphological/anatomical data for both living and fossil taxa.
Sections of this data can be difficult, or impossible, to collect for every taxon in the analysis.
For example, fossils rarely have molecular data and incomplete fossil preservation may reduce the number of anatomical characters available.
Additionally, it has become less common to collect anatomical characters for living taxa when molecular data is available (e.g. in \cite{slaterphylogenetic2013}, only 13\% of living taxa have coded anatomical characters).
Unfortunately this missing data can lead to errors in phylogenetic inference.
Simulations show that the ability of the Total Evidence method to recover the correct topology decreases when there is little overlap between coded anatomical characters in living and fossil taxa, and that the effect of missing data on topology is greatest when living taxa have few anatomical characters available \cite{GuillermeCooper}.
This is because (1) fossils will not be placed accurately within the correct clade if it contains no coded anatomical characters for living taxa; and (2) fossils have a higher probability of being placed within clades with more coded anatomical characters available for living taxa, regardless of whether this is the correct clade \cite{GuillermeCooper}.

The issues above highlight that it is crucial to have sufficient coded anatomical characters available for living taxa in a clade before using Total Evidence approaches.
However, it is unclear how many coded anatomical characters are actually available for living taxa, i.e. already coded from museum specimens and deposited in phylogenetic matrices accessible online, and how this data is distributed across clades.
Intuitively, most people assume this kind of data has already been collected, but empirical data suggest otherwise (e.g. in \cite{ronquista2012,slaterphylogenetic2013,beckancient2014}).
To investigate this further, we assess the number of available coded anatomical characters for living mammals to determine whether enough data exists to build reliable Total Evidence phylogenies.
We also determine whether the characters are phylogenetically overdispersed or clustered across mammalian orders.

%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------

\section{Materials and Methods}
\subsection{Data collection and standardisation}
We downloaded all cladistic matrices containing any living and/or fossil mammal taxa from three major public databases: MorphoBank (\url{morphobank.org} \cite{morphobank}), Graeme Lloyd's website (\url{graemetlloyd.com/matrmamm.html}) and Ross Mounce's GitHub repository (\url{github.com/rossmounce/cladistic-data}).
We also performed a systematic Google Scholar search for matrices that were not uploaded to these databases (see Electronic Supplementary Material (ESM) for details).
In total, we downloaded 286 matrices containing 5228 unique operational taxonomic units (OTUs). 
We used OTUs rather than species because entries in the matrices ranged from species to families. 
We standardised the taxonomy as described in the ESM and excluded OTUs that were not present in the phylogeny of \cite{BinindaEmonds} or the taxonomy of \cite{wilson2005mammal} to remove fossil species.
This resulted in 1601 unique OTUs from 286 matrices.

\subsection{Data availability and distribution}
To assess the availability of coded anatomical characters for each mammalian order and across mammals, we calculated the percentage of OTUs with cladistic data at three different taxonomic levels: family, genus and species.
We consider orders with \textless 25\% of living taxa with available anatomical characters as having low data coverage, and orders with \textgreater 75\% of living taxa with available anatomical characters as having high data coverage.

@@

%\input{Table_results}

@@

%\begin{figure}[!htbp]
%\centering
%    \includegraphics[width=1\textwidth]{figure1.pdf}
%\caption{Phylogenetic distribution of species with available coded anatomical characters across two orders (A: Primates; B: Carnivora).
%Blue branches indicate species with available coded anatomical characters.}
%\label{Figure_example_coverage}
%\end{figure}

@@

It is worth noting, however, that our analysis did not include all the matrices containing anatomical characters ever published.
In fact, our data collection procedure focused on including studies that provided matrices easily accessible, i.e. we did specifically not include any matrices that were only available in paper format (e.g. printed in books), non-reusable format (e.g. an image of the matrix)  or/and matrices available only upon request (e.g. by emailing the authors).
The difficulty of obtaining such matrices makes them less likely to be included in Total Evidence analyses, especially where the number of taxa to include in the phylogeny is high.

Despite the absence of good morphological/anatomical data coverage for living mammals, the Total Evidence method still seems to be the most promising way of combining living and fossil species in macroevolutionary analyses.
Following the recommendations in \cite{GuillermeCooper}, we must code anatomical characters for as many living species as possible. 
Fortunately, mammal specimens are usually readily available in natural history collections, therefore, we propose increased effort into coding anatomical characters from living species, possibly by engaging in collaborative data collection projects.
Such efforts would be valuable not only to phylogeneticists, but also to any researcher focusing understanding macroevolutionary patterns and processes.

%Biology letters various stuff
\section{Ethics statement}
N/A
\section{Data accessibility statement}
All data and code are available on GitHub (\url{https://github.com/TGuillerme/Missing_living_mammals}).
\section{Authors' Contributions}
TG and NC designed the study. TG analysed the data. TG and NC wrote the the manuscript.
\section{Competing Interests}
We have no competing interests.
\section{Acknowledgments}
We thank David Bapst, Graeme Lloyd, Nick Matzke, April Wright, Graham Slater, Peter Wagner and one anonymous reviewer.
\section{Funding statement}
This was funded by a European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant (proposal: 321696).
